---
editor_options: 
  markdown: 
    wrap: 72
---

# NFL Attendance Analysis

### By Evelyn Wake

# Introduction

## Problem Statement

### Introduction

Multiple times throughout a season, NFL organizations need to prepare
for game day. An important part of that is knowing how many fans will
attend a given game.

### Plan

There are several factors that could affect attendance, such as

-   A team's performance
-   The week number

The plan is to evaluate the relationship between the many aspects of a
team's performance and the fan attendance for that week, as well as the
week number vs fan attendance. This will be done by plotting different
variables pertaining to a team's performance and attendance for games.

# Packages Required

```{r}
# ggplot2 is required for more options while plotting graphs
library('ggplot2')

# gridExtra is required for orienting the display of multiple ggplot2 plots
library('gridExtra')
```

# Data Preparation

[Source](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-02-04/readme.md)

## Data Explanation

This data set is originally sourced from Pro Football Reference, a
reliable source for NFL statistics. The data was collected on February
4th, 2020

This data set contains 3 tables, which are described below:

### attendance.csv:

This table contains 8 variables pertaining to the attendance records for
each team on a week-by-week basis. Each row is the attendance record for
one team during one week

### standings.csv:

This table contains 15 variables pertaining to the overall standings of
the NFL on a season-by-season basis. Each row contains season statistics
for one team during a given season

### games.csv:

This table contains 19 variables, each representative of a different
statistic of a game. Each row is one singular game.

## Data Importing

```{r}
attendance = read.csv('data/attendance.csv')
games = read.csv('data/games.csv')
standings = read.csv('data/standings.csv')
```

## Data Cleaning

### Attendance

```{r}
# Convert the two columns containing the team's full name into one combined column
attendance$team_name = paste(attendance$team, attendance$team_name)

columns_of_interest = c('team_name', 'year', 'home', 'week', 'weekly_attendance')
# Narrow the table down to columns of interest
attendance = attendance[columns_of_interest]
# Delete rows where weekly attendance is NA to filter out bye weeks
attendance = attendance[!is.na(attendance$weekly_attendance), ]
```

Each of the columns selected serve a purpose in gathering valuable
attendance information

team_name: Utilized to specify which team the weekly attendance belongs
to

year: Utilized to specify which year the weekly attendance belongs to

home: Utilized to track the team's home attendance for that year

week: Utilized to specify which week the weekly attendance belongs to

weekly_attendance: Utilized to track the team's attendance for that week

### Games

```{r}
columns_of_interest = c('year', 'week', 'home_team', 'away_team', 'winner')
games = games[columns_of_interest]
#Only select regular season games, no playoffs
games = games[games$week %in% 1:17, ]
```

Each of the columns selected serve a purpose in gathering valuable game
information

year: Utilized to specify which year the game occurred in

week: Utilized to specify which week game occurred in

home_team: Utilized to specify the home team of that game

away_team: Utilized to specify the away team of that game

time: Utilized to specify the time that the game occurred at

winner: Utilized to specify which team won the game

### Standings

```{r}

standings$team_name = paste(standings$team, standings$team_name)
columns_of_interest = c('team_name', 'year', 'wins', 'simple_rating', 'strength_of_schedule', 'margin_of_victory', 'playoffs', 'sb_winner')
standings = standings[columns_of_interest]
#Convert 'playoffs' column to be numeric
standings$playoffs = factor(x=standings$playoffs, levels=c("Playoffs", "No Playoffs"))
levels(standings$playoffs) = c(1, 0)
#Convert 'sb_winner' column to be numeric
standings$sb_winner = factor(x=standings$sb_winner, levels=c("Won Superbowl", "No Superbowl"))
levels(standings$sb_winner) = c(1, 0)
```

Each of the columns selected serve a purpose in gathering valuable game
information

team_name: Utilized to specify the team that the standing belongs to

year: Utilized to specify which year the game occurred in

wins: Utilized to specify how many wins the team earned this season

simple_rating: Utilized to specify a team's relative performance

strength_of_schedule: Utilized to specify the strength of a team's
opponents

margin_of_victory: Utilized to specify how close the team's games were,
on average

playoffs: Utilized to specify if the team made the playoffs that year

sb_winner: Utilized to specify if the team won the superbowl that year

## Final Data Set

```{r}
head(attendance)
```

```{r}
head(games)
```

```{r}
head(standings)
```

# Exploratory Data Analysis

## Season Data vs Attendance

```{r}
# Select attendance records past 2000 so that they may be compared to the previous year
# Only select week 1, as that is the first game after the postseason
week_1_past_2000 = attendance[attendance$year > 2000 & attendance$week == 1, ]
# initiate vectors for season performance
simple_rating_vector = rep(0, nrow(week_1_past_2000))
strength_of_schedule_vector = rep(0, nrow(week_1_past_2000))
margin_of_victory_vector = rep(0, nrow(week_1_past_2000))
# Initiate vectors for postseason performance
won_sb_vector = rep(0, nrow(week_1_past_2000))
playoff_vector = rep(0, nrow(week_1_past_2000))
# Initiate vector to track the difference in attendance for current week 1 vs previous week 1
attendance_diff_vector = rep(0, nrow(week_1_past_2000))
# Iterate through each team's week in the dataframe
for(idx in 1:nrow(week_1_past_2000)){
  # Collect basic identifying information
  team_name = week_1_past_2000[idx,]$team_name
  winner = week_1_past_2000[idx,]$winner
  year = week_1_past_2000[idx,]$year
  # Get the team's strength of schedule for the current year
  strength_of_schedule = standings$strength_of_schedule[standings$year == year & standings$team_name == team_name]
  # Get the team's season performance for the previous year
  simple_rating = standings$simple_rating[standings$year == (year - 1) & standings$team_name == team_name]
  margin_of_victory = standings$margin_of_victory[standings$year == (year - 1) & standings$team_name == team_name]
  # Get the team's postseason performance for the previous year
  sb_result = standings$sb_winner[standings$year == (year - 1) & standings$team_name == team_name]
  playoff_result = standings$playoffs[standings$year == (year - 1) & standings$team_name == team_name]
  current_attendance = week_1_past_2000$weekly_attendance[week_1_past_2000$team_name == team_name & week_1_past_2000$year == year]
  # Calculate information on difference in attendance 
  last_year_attendance = attendance$weekly_attendance[attendance$team_name == team_name & attendance$year == (year - 1) & attendance$week == 1]
  attendance_diff_result = current_attendance - last_year_attendance
  # If team didn't exist the previous year, set all values to 0
  if(length(attendance_diff_result) == 0){
    simple_rating = 0
    margin_of_victory = 0
    sb_result = 0
    playoff_result = 0
    attendance_diff_result = 0
  }
  # Record the information in the vectors
  strength_of_schedule_vector[idx] = as.numeric(as.character(strength_of_schedule))
  simple_rating_vector[idx] = as.numeric(as.character(simple_rating))
  margin_of_victory_vector[idx] = as.numeric(as.character(margin_of_victory))
  won_sb_vector[idx] = as.numeric(as.character(sb_result))
  playoff_vector[idx] = as.numeric(as.character(playoff_result))
  attendance_diff_vector[idx] = attendance_diff_result
}
# Convert the vectors into a data frame and appropriately name the columns
season_df <- data.frame(strength_of_schedule_vector, simple_rating_vector, margin_of_victory_vector, won_sb_vector, playoff_vector, attendance_diff_vector)

names(season_df) = c("strength_of_schedule", "simple_rating", "margin_of_victory", "won_previous_sb", "made_previous_playoffs", "week_1_attendance_diff")
season_df$made_previous_playoffs = factor(x=season_df$made_previous_playoffs)
season_df$won_previous_sb = factor(x=season_df$won_previous_sb)
```

## Week-To-Week Performance

```{r}
# Function to parse a vector into a data frame for a team's performance during a week
parse_into_dataframe = function(v){
  df = as.data.frame(as.list(v))
  names(df) = c("year", "week", "home", "team_name", "won_game", "wins", "weeks_since_win")
  for(name in c("year", "week", "home", "won_game", "wins", "weeks_since_win")){
    df[name] = as.integer(df[name])
  }
  return(df)
}
# Initialize a data frame to collect data on a team's weekly performance
weekly_records = data.frame()
# Iterate through every game on record
for (idx in 1:nrow(games)){
  # Extract information on that week's game
  home_team = games[idx,]$home_team
  away_team = games[idx,]$away_team
  winner = games[idx,]$winner
  year = games[idx,]$year
  week = games[idx,]$week
  # Gather the pre-existing weekly records of both teams in the game
  home_records = weekly_records[weekly_records$team_name == home_team & weekly_records$year == year, ]
  away_records = weekly_records[weekly_records$team_name == away_team & weekly_records$year == year, ]
  # Initialize data based on if the team has played yet this year or not
  
  # Initialize number of wins for the home team
  if(length(home_records$team_name) == 0){
    home_win_count = 0
  }
  else{
    home_win_count = tail(home_records, n=1)$wins
  }
  # Initialize number of weeks since win for the home team
  if(length(home_records$team_name) == 0){
    home_weeks_since_win = 0
  }
  else{
    last_won_game = tail(home_records[home_records$won_game == 1, ], n = 1)$week
    if(length(last_won_game) == 0){
      home_weeks_since_win = as.integer(week)
    }
    else{
      home_weeks_since_win = as.integer(week) - last_won_game
    }
  }
  
  # Initialize number of wins for the away team
  if(length(away_records$team_name) == 0){
    away_win_count = 0
  }
  else{
    away_win_count = tail(away_records, n=1)$wins
  }
  # Initialize number of weeks since win for the away team
  if(length(away_records$team_name) == 0){
    away_weeks_since_win = 0
  }
  else{
    last_won_game = tail(away_records[away_records$won_game == 1, ], n = 1)$week
    if(length(last_won_game) == 0){
      away_weeks_since_win = as.integer(week)
    }
    else{
      away_weeks_since_win = as.integer(week) - last_won_game
    }
  }
  # Increment the win count of the winning team and specify each team's result
  if(winner == home_team){
    home_win_count = home_win_count + 1
    home_won = 1
    away_won = 0
  }
  else{
    away_win_count = away_win_count + 1
    home_won = 0
    away_won = 1
  }
  # Create the vector representation of each team's performance
  home_result = c(year, week, 1, home_team, home_won, home_win_count, home_weeks_since_win)
  away_result = c(year, week, 0, away_team, away_won, away_win_count, away_weeks_since_win)
  # Update the weekly records data frame
  if(nrow(weekly_records) == 0){
    weekly_records = parse_into_dataframe(home_result)
  }
  else{
    weekly_records = rbind(weekly_records, parse_into_dataframe(home_result))
  }
  weekly_records = rbind(weekly_records, parse_into_dataframe(away_result))
}
```

## Home Attendance Records

```{r}
# Narrow the data frame down to only the home teams for each week
home_records = weekly_records[weekly_records$home == 1, ] 
# Sort the home records data frame to match the attendance data frame format
home_records = home_records[order(home_records$team_name, home_records$year, home_records$week),]
#Initialize a vector for all of the attendance records corresponding to each home team 
attendance_recs = rep(0, nrow(home_records))
# Iterate through all of the home teams
for (idx in 1:nrow(home_records)){
  # Extract information on the home team, year, and week
  week = home_records[idx,]$week
  year = home_records[idx,]$year
  team = home_records[idx,]$team_name
  # Get the attendance data for that team during that week
  attendance_num = attendance$weekly_attendance[attendance$week == week & attendance$year == year & attendance$team_name == team]
  # Record the attendance data for that team during that week
  attendance_recs[idx] = attendance_num
  
}

# Add the attendance data to the data frame
home_records$attendance = attendance_recs

#Initialize a vector for all of the attendance records corresponding to each home team when differenced with that team's average home attendance for the year
attendance_diffs_recs = rep(0, nrow(home_records))
for (idx in 1:nrow(home_records)){
  # Extract information on the home team, year, and week
  week = home_records[idx,]$week
  year = home_records[idx,]$year
  team = home_records[idx,]$team_name
  # Get the attendance data for that team during that week
  attendance_num = home_records$attendance[home_records$week == week & home_records$year == year & home_records$team_name == team]
  year_attendance = home_records$attendance[home_records$year == year & home_records$team_name == team]
  average_attendance = mean(year_attendance)
  # Record the attendance difference data for that team during that week
  attendance_diffs_recs[idx] = attendance_num - average_attendance
  
}

# Add the attendance difference data to the data frame
home_records$attendance_diff = attendance_diffs_recs
```

# Plots & Analysis

## Number Of Wins

```{r}
# Plot data comparing number of wins and attendance data
dev.new(width=9, height=4.5)
layout_mtx = matrix(c(1, 2), 1, 2)
lay.mat = layout_mtx
layout(mat=lay.mat)   
plot(home_records$wins, home_records$attendance / 1000, xlab="Number of Wins", ylab = "Game Attendance (Thousands)", main = "Wins vs Attendance")
box(which="figure", col='gray')  
plot(home_records$wins, home_records$attendance_diff, xlab = "Number of Wins", ylab = "Game Attendance - Season Average Attendance", main = "Wins vs Attendance Difference")
box(which="figure", col='gray')  
```

### Analysis

As shown in the two scatter plots above, there does not appear to be
much correlation between a team's number of wins and their overall
attendance for that season. Although there is a lot of variance within
the data, that is likely explained by other factors such as location.
The variance appears to lessen as the number of wins increases, but this
is to be expected as it is rarer for a team to get a higher number of
wins.

## Weeks Since Win

```{r}
# Plot data comparing weeks since win and attendance data
dev.new(width=12, height=4.5)
layout_mtx = matrix(c(1, 2), 1, 2)
lay.mat = layout_mtx
layout(mat=lay.mat)   
plot(home_records$weeks_since_win, home_records$attendance / 1000, xlab="Weeks Since Win", ylab = "Game Attendance (Thousands)", main = "Weeks Since Win vs Attendance")
box(which="figure", col='gray')  
plot(home_records$weeks_since_win, home_records$attendance_diff, xlab = "Weeks Since Win", ylab = "Game Attendance - Season Average Attendance", main = "Weeks Since Win vs\n Attendance Difference")
box(which="figure", col='gray')  
```

### Analysis

Similarly to the previous scatter plots, there is not much correlation
between attendance records and the number of weeks since a win. This is
unsurprising as it supports the factor that within a season, the overall
number of wins does not affect attendance in a major way.

## Week Number

```{r}
# Plot data comparing week number and attendance data
dev.new(width=9, height=4.5)
layout_mtx = matrix(c(1, 2), 1, 2)
lay.mat = layout_mtx
layout(mat=lay.mat)    
gg1 = ggplot(home_records, aes(x=week, y=attendance / 1000)) + 
  geom_bar(stat = "identity") + xlab("Week Number") +
  ylab("Attendance (Thousands)") + ggtitle("Week Number vs Attendance (Thousands)")
gg2 = ggplot(home_records, aes(x = week, y = attendance_diff)) +
    geom_point() + xlab("Week Number") +
  ylab("Avg Attendance - Season Avg")  + ggtitle("Week Number vs Attendance Difference")
grid.arrange(gg1, gg2)
```

### Analysis

As shown by the bar plot, the overall weekly attendance dips down during
the middle of the season, and then rises back up by week 12. This is
likely explained by the fact that the middle of the season is the least
exciting compared to the beginning and the end of the season. In the
beginning, fans are excited for the return of football, and are more
likely to go to games. Towards the end of the season, competition rises
as teams begin fighting harder for playoff spots, and games are likely
to become more interesting.

However, when looking at the scatter plot comparing week number and
attendance difference, something interesting appears.There are more
games that are below a team's average attendance for that season. This
could possibly be explained by the fact that towards the end of the
season, a team's postseason fate becomes more clear. Specifically around
week 13, the amount of games with relatively low attendance increases.
By that time, a team could potentially be 2-10, with it being extremely
unlikely that they will attend the playoffs. On the other hand, a team
could be 12-0, and they will almost certainly make the playoffs that
year. With the postseason implications of those scenarios being
relatively low, teams might not try as hard to win, and even bench some
of their star players as to not risk injury. With that being said, those
games might not be as interesting to fans, and they might be less likely
to attend.

## Season Data

```{r}
# Plot data comparing if the team made the previous playoffs and attendance data
ggplot(season_df, aes(x=made_previous_playoffs, y=week_1_attendance_diff)) + 
    geom_boxplot() + xlab("Made Previous Playoffs") + ylab("Week 1 Attendance - Previous Week 1 Attendance") + scale_x_discrete(labels = c("No", "Yes"))  + ggtitle("Previously Made Playoffs and Week 1 Attendance Difference")

```

### Analysis

Although the median week 1 difference in attendance between teams who
have made the playoffs the previous season and those who hadn't are
nearly identical, there exists an interesting difference between the two
boxplots. Both the 25th and 75th percentile for attendance for teams who
had previously made playoffs are higher.

Although not influential on the analysis itself, it is interesting to
see a -50000 difference between one team's week 1 attendance.

```{r}
# Plot data comparing if the team won the previous superbowl and attendance data
ggplot(season_df, aes(x=won_previous_sb, y=week_1_attendance_diff)) + 
    geom_boxplot(alpha=0.5) + xlab("Won Previous Superbowl") + ylab("Week 1 Attendance - Previous Week 1 Attendance") + scale_x_discrete(labels = c("No", "Yes")) + ggtitle("Previous Superbowl Result and Week 1 Attendance Difference")


```

### Analysis

The median week 1 difference in attendance between teams who have won
the Superbowl the previous season and those who hadn't are quite close.
The plot for teams that have previously won the Superbowl appears to be
a condensed version of the teams that hadn't, which makes sense given
the relatively small sample size of teams that have won the Superbowl.

```{r}
# Plot data comparing regular season information and week 1 attendance difference
dev.new(width=13.5, height=4.5)
layout_mtx = matrix(c(1, 2, 3), 1, 3)
lay.mat = layout_mtx
layout(mat=lay.mat)   
plot(season_df$strength_of_schedule, season_df$week_1_attendance_diff, xlab = "Strength of Schedule", ylab = "Week 1 Attendance - Previous Week 1 Attendance", main = "Strength of Schedule\n vs Week 1 Attendance Diff")
box(which="figure", col='gray')  
plot(season_df$simple_rating, season_df$week_1_attendance_diff, xlab = "Previous Season's Simple Rating", ylab = "Week 1 Attendance - Previous Week 1 Attendance", main = "Previous Season's Simple Rating\n vs Week 1 Attendance Diff")
box(which="figure", col='gray')  
plot(season_df$margin_of_victory, season_df$week_1_attendance_diff, xlab = "Previous Season's Margin of Victory", ylab = "Week 1 Attendance - Previous Week 1 Attendance", main = "Previous Season's Margin of\n Victory vs Week 1 Attendance\n Diff")
box(which="figure", col='gray')  
```

### Analysis

Interestingly, it appears that a team's performance the previous season,
as well as the strength of their schedule does not have much impact on
their week 1 attendance when compared to the previous year.

# Summary

## Problem Statement

To reiterate the problem statement:

"Multiple times throughout a season, NFL organizations need to prepare
for game day. An important part of that is knowing how many fans will
attend a given game."

## Addressing the Problem Statement

To address the problem statement, I utilized the NFL attendance data set
to extract multiple features of a team's performance, and compared it to
attendance data for both weeks and seasons. I also compared attendance
data to itself, and utilized that difference to illustrate any impact a
team's performance might have on the difference.

Additionally, I compared attendance data to week numbers to see how
attendance data changes throughout a season

## Summarizing Insights

Oddly enough, it did not appear that there was much correlation between
a team's performance in-season and their attendance for that season.
However, if a team had made the playoffs during the previous season, the
week 1 attendance for the next season is likely to be higher than the
week 1 attendance of the previous season

One factor that did appear to impact a team's attendance is week number.
It was found that towards the middle of the season, attendance appears
to dip down, but rise back up around week 12.

## Consumer Implications

When preparing for a game, teams should largely pay attention to what
week in the season it is, and compare that to previous weeks in the
season. If it is week 1, the team should pay attention to how many fans
attended the week 1 game last season.

They should also factor in if the teams made the playoffs the previous
year.

Factoring in the aforementioned details could provide an idea of how to
adequately prepare for upcoming games when considering attendance.

## Limitations

One of the major limitations of this analysis was that I was only
working in 2 dimensions at a time. It is possible that two or more
factors, when combined, uncover a correlation that wouldn't be found by
either alone

Another major limitations lies within the data. It is important to
consider the weather occurring during a game. Fans are more likely to
attend a game on a sunny day compared to a game when it is snowing or
raining.

Alongside this, location could be incredibly important to consider. For
example, it is more likely to snow during a home game for a team that is
further north than a team that is further south. Nearby population
density could also play a huge role in this.

Finally, it would be important to consider each stadium's capacity. Some
teams might have lower attendance because their stadium simply cannot
hold as many fans. I attempted to account for this by utilizing average
attendance, but I believe it would be better to utilize the capacity
percentage instead of number of attending fans
